Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-08 paper.tex
alg updated

The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
and the node is expanded.
The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
is guided using moving virtual goal~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
are generated with probability $\gb$. 
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

@@

The core of the method is the expansion procedure, which generates new collision-free nodes around $\qnear\in\T$.
The algorithm is listed in Alg.~\ref{alg::expand}.
For each conformation $l \in \L$, the expansion generates up to $m$ samples $q$ with maximal radii around $\qnear$.
The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
parameter.
The rotation part of $q$ is generated using $U(0,2\pi)$.
The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
with $s$, the next lower $s$ is used.
The expansion step is repeated for each conformation $l \in L$.

todo atom metric

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
%    $newnodes = $ false\;
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        $\qnew = \emptyset$; // empty configuration\\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$) {\bf and} \\ $(\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear))$} {
                $\qnew = q$\;
            } 
        }
        $\T$.addNode($q$)\;
        $\T$.addEdge($\qnear,q$)\;
    }
}
%\return $S$\;
\end{algorithm}

>>>>

The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
is guided using moving virtual goal~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
are generated with probability $\gb$. 
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

@@

The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with maximal
scale.
For a given conformation $l \in \L$, first the maximal scale $\smax$ is considered and
$m$ samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.

The sampling-based methods are sensitive to the employed metric, especially if the objects are not symmetrical, which is the
case of the non-spherical ligands.
To consider the actual shape of the ligand (which is different in each conformation) and to allow finding such configurations
that approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
        Alg.~\ref{alg::expand}).

The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
parameter.
The rotation part of $q$ is generated using $U(0,2\pi)$.
The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
with $s$, the next lower $s$ is used.
The expansion step is repeated for each conformation $l \in L$.

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        $\qnew = \emptyset$; // empty configuration\\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to next conformation
        }
    }
}
%\return $S$\;
\end{algorithm}

The configruation tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
one new configuration can be added.
The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
The collision detection can be evaluated fast using hiearchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
